The Units of Permissionless Consensus: \\ Towards Mobile and Edge Computing

### 2022-11-06 theoretical-consensus/article/article.tex
add page 07

The advancements of the internet 
more than potentiated the revolution and what we now call the permissionless consensus problem was 
finally born. Without forgetting the previous attempts, the first practical permissionless consensus algorithm was 
proposed by Nakamoto in \cite{nakamoto2008bitcoin}. It is a proof-of-work consensus protocol that resembles a "replicated 
state machine" where the independent participants reach agreement not only about transactional values, 
but also about their order - naturally forming the underlying structure of what is now known as a blockchain. 
"Proof-of-work is essentially one-CPU-one-vote" and this is the novelty 
introduced by Bitcoin \cite{pass2016hybrid,pass2016hybrid2}. The focus shifted 
for decentralized systems and after proof-of-work many other consensus mechanisms have been proposed, 
based on different consensus units, like proof-of-stake, proof-of-space, proof-of-authority, etc. The chaotic 
diversity of new consensus protocols gave also room for endless reviews, overviews and comparisons \cite{survey-dist-consensus, BAMAKAN2020113385, 8400278, natoli2019deconstructing, 8629877, 9376868, BOURAGA2021114384}. 
The authors of these surveys often put multiple dimensions into comparison, like fault tolerance, scalability, or energy consumption,
and, among those, some focused their efforts on mechanisms that may work in resource-constrained networks \cite{9104713, 8661654, 8168250, SALIMITARI2020100212}. 
This paper will try to identify the common conclusions from these comparisons, while looking into the state-of-the-art and novel approaches for running permissionless
consensus protocols in mobile and edge devices.

@@

>>>>

The advancements of the internet 
more than potentiated the revolution and what we now call the permissionless consensus problem was 
finally born. Without forgetting the previous attempts, the first practical permissionless consensus algorithm was 
proposed by Nakamoto in \cite{nakamoto2008bitcoin}. It is a proof-of-work consensus protocol that resembles a "replicated 
state machine" where the independent participants reach agreement not only about transactional values, 
but also about their order - naturally forming the underlying structure of what is now known as a blockchain. 
"Proof-of-work is essentially one-CPU-one-vote" and this is the novelty 
introduced by Bitcoin \cite{pass2016hybrid,pass2016hybrid2}. The focus shifted 
for decentralized systems and after proof-of-work many other consensus mechanisms have been proposed, 
based on different consensus units, like proof-of-stake, proof-of-space, proof-of-authority, etc. The chaotic 
diversity of new consensus protocols gave also room for endless reviews, overviews and comparisons \cite{survey-dist-consensus, BAMAKAN2020113385, 8400278, natoli2019deconstructing, 8629877, 9376868, BOURAGA2021114384}. 
The authors of these surveys often put multiple dimensions into comparison, like fault tolerance, scalability, or energy consumption,
and, among those, some focused their efforts on mechanisms that may work in resource-constrained networks \cite{9104713, 8168250, queralta2021blockchain, SALIMITARI2020100212}. 
This paper will try to identify the common conclusions from these comparisons, while looking into the state-of-the-art and novel approaches for running permissionless
consensus protocols in mobile and edge devices.

@@

\section{Resource-Constrained Networks}

\subsection{Characteristics}

Resource-constrained networks are a class of networks that are characterized by the
limited access to resources, such as energy, bandwidth, storage, etc.
These networks are typically seen in IoT applications, where the devices are
usually low-powered, with limited computational resources, but are expected to 
collectively operate with high availability, at least the network as a whole, 
preferably without the need for frequent maintenance.
These networks are also expected to be dynamic, with frequent changes in the network
topology, by the unpredictable addition and removal of nodes. 
The nodes are also expected to be heterogeneous, with different capabilities, and may be
connected to the network through different communication channels. 
Mobility may also be a feature, as the nodes may spatially move around, 
and may directly pair with different nodes at different moments.
The nodes may also be individually unreliable, untrusted, uncoordinated, unmanageable, may fail at any time, 
and may be disconnected from the rest of the network.

Nowadays, the typical resource-constrained networks follow a centralized architecture
with a client-server model, where the network is controlled by a single entity,
which is responsible for the management of the nodes, including authentication,
authorization, and communication. This single-point-of-failure is also a major
scalability bottleneck, and a maintenance burden, as the network grows in number of devices \cite{SALIMITARI2020100212}.

\subsection{Challenges}

Having in mind these characteristics of resource-constrained networks,
and the drawbacks of the centralized architecture, it may be postulated the 
need for a decentralized or permissionless shift that addresses some the most important challenges
of these deployments \cite{SALIMITARI2020100212}:

\begin{itemize}
  \item \emph{Scalability}:  in terms of the number of available devices - decentralizing may allow
  for better, faster and less costly maintenance of the network, assuring more availability and reliability
  and cutting the single-entity dependency;
  \item \emph{Security}: in terms of the network as a whole - removing the single-point-of-failure 
  may reduce the attack surface, and may also allow or require more flexible and dynamic security policies;
  \item \emph{Privacy}: in terms of the network data exchange features - decentralizing may allow for
  a user-centric approach, where the users are in control of the data that is shared with the network; 
\end{itemize}

Permissionless consensus is a promising solution for these challenges, as an enabler for
decentralized environments. The use cases for these networks are diverse. Nevertheless,
as long as consensus is achieved, the network as a whole can be abstractly seen and used
as a trusted entity that runs as a trustless decentralized service. Applications may then
be built on top of it, and the network may be used as a platform for other services \cite{queralta2021blockchain}.
However, there are some key aspects that need to be addressed
in order to make this approach viable for resource-constrained networks, as we saw in the previous section:

\begin{itemize}
  \item \emph{Resource consumption}: the consensus protocols should be resource-efficient, for example, in terms of energy, bandwidth, or storage, 
  as the nodes are expected to be low-powered devices, with limited computational, communication, and storage resources;
  \item \emph{Heterogeneity}: the consensus protocols should account for different device characteristics, 
  as the nodes are expected to be heterogeneous in terms of their capabilities;
  \item \emph{Fault tolerance}: the consensus protocols should be fault-tolerant, as the nodes are expected to be unreliable, 
  untrusted, uncoordinated, or unmanageable;
  \item \emph{Openness}: the consensus protocols should be open and trustless, as the nodes are expected to be 
  dynamically added and removed from the network;
  \item \emph{Security}: the consensus protocols should be resistant to a multitude of new attacks,
   especially if it turns out to be easy for malicious entities to overcome resource constraints;
\end{itemize}

The overall theoretical properties of the consensus protocols still need to be taken into account,
as the goal is to have a proved functional mechanism that achieves Termination, Agreement, Validity, and Integrity,
while playing with the trade-offs imposed by the trilemma (See Fig. \ref{fig:trilemma}) \cite{survey-dist-consensus}.

\subsection{Related Work}

There have been several attempts to address the challenges of resource-constrained networks,
and to understand the limitations of the existing consensus protocols in this context.
A common conclusion is that the existing protocols are not suitable for these networks,
due either to the high resource consumption, or to the centralization of various aspects of the protocol,
or also to high communication overheads, that are translated to high latency and low throughput \cite{SALIMITARI2020100212},
making these protocols unsuitable for resource-constrained applications. Other concerns are the
security issues, as some attempts to overcome the resource constraints may lead to new vulnerabilities,
which may be exploited by malicious and more powerful entities.

A thoroughly mentioned approach is the IOTA Tangle protocol \cite{popov2018tangle}, which
presents a friendly design for resource-constrained networks, infinitely scalable, and with low latency.
However, there are still known security and centralization limitations, as presented in the sections before.
Another concept that has been explored is sharding \cite{SALIMITARI2020100212, queralta2021blockchain}, which is a
technique that allows for the partitioning of the network into smaller sub-networks,
that are then managed by different entities, validating transactions in parallel. 
This approach may be used to overcome the scalability bottleneck,
but it may also introduce new issues that are still to be researched.
Frankly, there are indeed some well established protocols that, all in all, can be adapted to networks with
resource-constrained devices that are not necessarily short on the resources needed to run the protocols.
For example, a lot of \emph{PoS} related protocols may be adapted to resource-constrained networks, where 
the stake is the only considerable resource that the nodes need. There are a multitude of projects that are
currently exploring this approach, for example, with the Ethereum transition to \emph{PoS} \cite{buterin2017casper}.
However, this does not address the hipothesis of taking advantage of the very nature of these limited environments, 
as simply choosing a BFT protocol that is resource-efficient
may also put us far apart from the ultimate decentralization and scalability goal of 
other and fairer means of consensus participation.
An interesting idea is based on the use of \emph{Proof of Location} for 
coordinating the consensus. The easiness of movement and targetted location deployment
of a lot of these low-power devices may be used to achieve a decentralized consensus
protocol, where elected nodes may be somehow chosen based on their location, 
either by proximity, or by some other withnessing mechanism. Variations and 
extensions of this idea, with different and wide use cases, have been seen in
\cite{foam2018location, xyo2018location}

\section{Conclusion}

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Donec auctor, nisl eget aliquam aliquam.

\section{Acknowledgements}

This work was supported by the 2022 XRP Ledger Trust Scholarship Program for Master's Students
at the University of Tartu.
